Memory-Efficient Mixed-Precision Implementations\\for Robust Explicit Model Predictive Control

### 2019-07-17 emsoft-2019/camera_ready/Cont_Syn.tex
colors-->blue

Unlike the initial $\Delta$, $\varepsilon_Q$ needs to be strictly positive (line 4)
for mixed-precision tuning tools to work correctly. \Mahmoud{Typical values for $\varepsilon_Q$ can range from $10^{-4}$ up-to $10^{-2}$. Increasing the $\varepsilon_Q$ leads into more memory save over $H$ and $K$, while requiring more precision for $F$ and $G$. Of course, too large values for $\varepsilon_Q$ can result in failure of RMPC design. However, this is usually not the case over reasonable choices for $\varepsilon_Q$.}
Line 5 computes the width of the tubes $\varepsilon$ as shown in~\autoref{eq:epsilontot},
which will be used to bound the error due to selecting the wrong region
($\varepsilon_{A/D}$ is a constant parameter dependent on the specifics of the
converter).

@@

If the region error is smaller than $\Delta$, we still have an error budget \texttt{errorAct} left
over to implement the control actions in finite precision (line 11).
For the precision assignment, we first call Daisy with the expressions for the
control actions given by $F$ and $G$, and the remaining error
budget as the error bound (line 12).
Then we call Daisy with the expressions of the region bounds given by $H$
and $K$, now with the error bound $\varepsilon_Q$ (line 13).\Mahmoud{
\begin{remark}
From a theoretical point of view, the error due to wrong region selection does not necessarily converge as each new designed controller may differ from the previous ones. Therefore, it is not easy to give an upper bound over the number of times that the robust MPC design needs to be repeated. However, by rewriting \autoref{eq:RMPC_prob}, we have
\begin{align*}
\label{eq:argmin}
&u_0(x_0,\Delta),\cdots,u_{N-1}(x_0,\Delta)=\\
&argmin \max_{w_0,\cdots,w_{N-1}} \sum_{i=0}^{N-1}(x_i^TQx_i+u_i^TRu_i) + x_N^TQ_Fx_N\nonumber\\
&\text{s.t.} \quad x_{i+1}=Ax_i+Bu_i + E w_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
%\quad &u_i=\mu x_i+v_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&u_i\in\mathcal{U},x_i\in\mathcal{X},\quad \forall w_i\in\mathcal{W}_{\Delta},\,\,\forall i\in\{0,1,\cdots,N\},
\end{align*}
one can use results on continuity of $argmin$ function, assuming its unique solution to every $\Delta$, to  compute the Lipschitz constant for it over variations of $\Delta$. The computed Lipschitz constant will be only dependent on the dynamics of the system depicted in \autoref{eq:DSS} and hence can be used to evaluate the maximum error over incorrect region selection. This Lipschitz constant can be quite large, leading into a very conservative bound on maximum number of iterations before convergence. However, from an experimental point of view, using a reasonable $\varepsilon_{SAFE}$, our observations show that convergence always happened after only a few iterations.
\end{remark}}
In both cases, Daisy determines a quantization of the input matrices (Daisy also
determines the fixed-point formats of the arithmetic operations, which we ignore here).
Daisy either computes a minimal uniform fixed-point precision (i.e. word length),
or it returns a mixed-precision assignment, where each value and
intermediate operation can potentially have a different word length.
To determine the minimal uniform precision, Daisy performs a linear search.
It starts from the smallest precision (1 bit) and checks whether it satisfies the
error bound. If not, the precision is increased by 1 bit and the first precision
that satisfies the error bound is returned.
Minimal uniform precision is used as a starting point for mixed-precision tuning,
which attempts to assign a lower precision to some variables. 
In our experiments in~\autoref{sec:experiments}, we compare the two modes and show
that mixed precision leads to smaller memory footprints.

Note that Daisy always returns a quantization (at least for our linear
expressions). If the error bound given is very small (but larger than zero),
Daisy will return a fixed-point precision with a large number bits, but will not
fail.\\
\Mahmoud{In \cite{Alessio2009} it is shown that the upper bound over the number of regions $P$ is $2^q$, where $q$ denotes the total number of constraints in \autoref{eq:RMPC_prob}. In theory one cannot find an upper bound for the maximum number of hyperplanes enclosing a region. To compute the incorrect region selection error, assuming that each partition at most has $\bar n_f$ faces, in the worst case $i\times\bar n_f\times 2^q$ affine computations would be enough, where $i$ denotes the number of times that RMPC design is repeated for updated $\Delta$s. In our experiments, computation of incorrect region selection takes only a few minutes for our largest tested benchmarks. %\underline{\textbf{Rocco: What about complexity of Daisy's computations in terms of number of regions $P$ and $n$?}} 		
\\
The complexity of precision tuning is linear with respect to the number of regions and hyperplanes. So, $P+\bar n_f\times P$ is an upper bound for the number of problems we need to encode in Daisy.
In each encoding, the uniform-precision detection takes linear number of steps before convergence. It starts from 1-bit and increases the precision one by one until the first valid configuration. The complexity of mixed-precision tuning, with delta-debugging, is $n\times log(n)$ in average, and bounded by $n^2$ in the worst case, with n number of variables.}

>>>>

Unlike the initial $\Delta$, $\varepsilon_Q$ needs to be strictly positive (line 4)
for mixed-precision tuning tools to work correctly. \Mahmoudd{Typical values for $\varepsilon_Q$ can range from $10^{-4}$ up-to $10^{-2}$. Increasing the $\varepsilon_Q$ leads into more memory save over $H$ and $K$, while requiring more precision for $F$ and $G$. Of course, too large values for $\varepsilon_Q$ can result in failure of RMPC design. However, this is usually not the case over reasonable choices for $\varepsilon_Q$.}
Line 5 computes the width of the tubes $\varepsilon$ as shown in~\autoref{eq:epsilontot},
which will be used to bound the error due to selecting the wrong region
($\varepsilon_{A/D}$ is a constant parameter dependent on the specifics of the
converter).

@@

If the region error is smaller than $\Delta$, we still have an error budget \texttt{errorAct} left
over to implement the control actions in finite precision (line 11).
For the precision assignment, we first call Daisy with the expressions for the
control actions given by $F$ and $G$, and the remaining error
budget as the error bound (line 12).
Then we call Daisy with the expressions of the region bounds given by $H$
and $K$, now with the error bound $\varepsilon_Q$ (line 13).\Mahmoudd{
\begin{remark}
From a theoretical point of view, the error due to wrong region selection does not necessarily converge as each new designed controller may differ from the previous ones. Therefore, it is not easy to give an upper bound over the number of times that the robust MPC design needs to be repeated. However, by rewriting \autoref{eq:RMPC_prob}, we have
\begin{align*}
\label{eq:argmin}
&u_0(x_0,\Delta),\cdots,u_{N-1}(x_0,\Delta)=\\
&argmin \max_{w_0,\cdots,w_{N-1}} \sum_{i=0}^{N-1}(x_i^TQx_i+u_i^TRu_i) + x_N^TQ_Fx_N\nonumber\\
&\text{s.t.} \quad x_{i+1}=Ax_i+Bu_i + E w_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
%\quad &u_i=\mu x_i+v_i, \quad\forall i\in\{0,1,\cdots,N-1\}\nonumber\\
&u_i\in\mathcal{U},x_i\in\mathcal{X},\quad \forall w_i\in\mathcal{W}_{\Delta},\,\,\forall i\in\{0,1,\cdots,N\},
\end{align*}
one can use results on continuity of $argmin$ function, assuming its unique solution to every $\Delta$, to  compute the Lipschitz constant for it over variations of $\Delta$. The computed Lipschitz constant will be only dependent on the dynamics of the system depicted in \autoref{eq:DSS} and hence can be used to evaluate the maximum error over incorrect region selection. This Lipschitz constant can be quite large, leading into a very conservative bound on maximum number of iterations before convergence. However, from an experimental point of view, using a reasonable $\varepsilon_{SAFE}$, our observations show that convergence always happened after only a few iterations.
\end{remark}}
In both cases, Daisy determines a quantization of the input matrices (Daisy also
determines the fixed-point formats of the arithmetic operations, which we ignore here).
Daisy either computes a minimal uniform fixed-point precision (i.e. word length),
or it returns a mixed-precision assignment, where each value and
intermediate operation can potentially have a different word length.
To determine the minimal uniform precision, Daisy performs a linear search.
It starts from the smallest precision (1 bit) and checks whether it satisfies the
error bound. If not, the precision is increased by 1 bit and the first precision
that satisfies the error bound is returned.
Minimal uniform precision is used as a starting point for mixed-precision tuning,
which attempts to assign a lower precision to some variables. 
In our experiments in~\autoref{sec:experiments}, we compare the two modes and show
that mixed precision leads to smaller memory footprints.

Note that Daisy always returns a quantization (at least for our linear
expressions). If the error bound given is very small (but larger than zero),
Daisy will return a fixed-point precision with a large number bits, but will not
fail.\\
\Mahmoudd{In \cite{Alessio2009} it is shown that the upper bound over the number of regions $P$ is $2^q$, where $q$ denotes the total number of constraints in \autoref{eq:RMPC_prob}. In theory one cannot find an upper bound for the maximum number of hyperplanes enclosing a region. To compute the incorrect region selection error, assuming that each partition at most has $\bar n_f$ faces, in the worst case $i\times\bar n_f\times 2^q$ affine computations would be enough, where $i$ denotes the number of times that RMPC design is repeated for updated $\Delta$s. In our experiments, computation of incorrect region selection takes only a few minutes for our largest tested benchmarks. %\underline{\textbf{Rocco: What about complexity of Daisy's computations in terms of number of regions $P$ and $n$?}} 		
\\
The complexity of precision tuning is linear with respect to the number of regions and hyperplanes. So, $P+\bar n_f\times P$ is an upper bound for the number of problems we need to encode in Daisy.
In each encoding, the uniform-precision detection takes linear number of steps before convergence. It starts from 1-bit and increases the precision one by one until the first valid configuration. The complexity of mixed-precision tuning, with delta-debugging, is $n\times log(n)$ in average, and bounded by $n^2$ in the worst case, with n number of variables.}

### 2019-07-17 emsoft-2019/camera_ready/Error_Analysis.tex
colors-->blue

% Note that $\mathcal R_i\subset \bar{\mathcal R}_i$. In other words, $\bar{\mathcal R}_i$ is the expanded version of $\mathcal R_i$.
% Note that $\bar{\mathcal R}_i\cap \bar{\mathcal R}_j$ gives us the states where one of the two regions $\mathcal R_i, \mathcal R_i$ might be selected instead of the other one because of the fixed-precision implementation of the regions.   
% $\bar{\mathcal R}_i$ is the expanded version of $\mathcal R_i$ that include all states in $\mathcal R_i$this definition expands the regions $\mathcal $
 %That is, with infinite-precision arithmetic
 Therefore, the first term in~\autoref{eq:delta} can be computed less conservatively by maximizing only over 
 the expanded borders:
 \begin{equation}
 \label{eq:maximization}
 \max_{x,i,j}\{\|F_ix+G_i-F_jx-G_j\|,\,x\in \bar{\mathcal B}_{ij}\}.
% \max_{\forall i,j.\mathit{neighbor}(\mathcal{R}_{i},\mathcal{R}_{j})}&|F_{i}\statevar+G_{i} - (F_{j}\statevar+G_{j})|
 \end{equation}\Mahmoud{ 
 \begin{remark}
 	Our formulation is general enough to take the effect of measurement/process noise into account. More specifically, $\Delta$ in $\mathcal{W}_{\Delta}$ can in general give upper-bound over the sum of uncertainties with different sources. However, this is not subject of study of the current paper. Moreover, to cover nonlinear systems, one approach is to linearize them around an appropriately selected equilibrium point of the system and to model the differences as a bounded disturbance as in \autoref{eq:RMPC_prob}. Taking the upper bound over nonlinearities into account will result in enlarging $\Delta$ which characterizes $\mathcal{W}_{\Delta}$. For reasonable values of $\Delta$, in principle RMPC design should work successfully. Further, it should be noticed that the rest of analysis (over MATLAB and Daisy) do not rely on the LTI property of the under study system and could be adjusted for other families of controllers. However, such analysis are beyond the scope of this paper as (perturbed) LTI systems are considered as an important class of dynamical systems.
 	\end{remark}}

@@

While our actual synthesis algorithm (\autoref{sec:Controller_Synthesis}) does
not compute the errors $\varepsilon_Q$ and $err(v_{i})_{p}$ explicitly, the
error verification is used internally by Daisy to determine a suitable precision
$p$ and can also be called explicitly without the precision optimization.
\Mahmoud{\begin{remark}
	MATLAB and Daisy are solely used for offline computations. The output of our algorithm contains (almost) all the information required for implementation of the verified controller over embedded systems. Therefore, the same savings can be expected when the controller is implemented on an actual embedded system.
\end{remark}}
\begin{comment}
\textcolor{red}{Sadegh: This is long comment on the content of this section. I cannot figure out how this line of reasoning is mapped to the implementation. We can discard my comment given that we are close to the deadline.}

>>>>

% Note that $\mathcal R_i\subset \bar{\mathcal R}_i$. In other words, $\bar{\mathcal R}_i$ is the expanded version of $\mathcal R_i$.
% Note that $\bar{\mathcal R}_i\cap \bar{\mathcal R}_j$ gives us the states where one of the two regions $\mathcal R_i, \mathcal R_i$ might be selected instead of the other one because of the fixed-precision implementation of the regions.   
% $\bar{\mathcal R}_i$ is the expanded version of $\mathcal R_i$ that include all states in $\mathcal R_i$this definition expands the regions $\mathcal $
 %That is, with infinite-precision arithmetic
 Therefore, the first term in~\autoref{eq:delta} can be computed less conservatively by maximizing only over 
 the expanded borders:
 \begin{equation}
 \label{eq:maximization}
 \max_{x,i,j}\{\|F_ix+G_i-F_jx-G_j\|,\,x\in \bar{\mathcal B}_{ij}\}.
% \max_{\forall i,j.\mathit{neighbor}(\mathcal{R}_{i},\mathcal{R}_{j})}&|F_{i}\statevar+G_{i} - (F_{j}\statevar+G_{j})|
 \end{equation}\Mahmoudd{ 
 \begin{remark}
 	Our formulation is general enough to take the effect of measurement/process noise into account. More specifically, $\Delta$ in $\mathcal{W}_{\Delta}$ can in general give upper-bound over the sum of uncertainties with different sources. However, this is not subject of study of the current paper. Moreover, to cover nonlinear systems, one approach is to linearize them around an appropriately selected equilibrium point of the system and to model the differences as a bounded disturbance as in \autoref{eq:RMPC_prob}. Taking the upper bound over nonlinearities into account will result in enlarging $\Delta$ which characterizes $\mathcal{W}_{\Delta}$. For reasonable values of $\Delta$, in principle RMPC design should work successfully. Further, it should be noticed that the rest of analysis (over MATLAB and Daisy) do not rely on the LTI property of the under study system and could be adjusted for other families of controllers. However, such analysis are beyond the scope of this paper as (perturbed) LTI systems are considered as an important class of dynamical systems.
 	\end{remark}}

@@

While our actual synthesis algorithm (\autoref{sec:Controller_Synthesis}) does
not compute the errors $\varepsilon_Q$ and $err(v_{i})_{p}$ explicitly, the
error verification is used internally by Daisy to determine a suitable precision
$p$ and can also be called explicitly without the precision optimization. \Mahmoudd{To encourage reproducible research, our implementation is made available. \footnote{Please insert the link to our package here!!!}}
\Mahmoudd{\begin{remark}
	MATLAB and Daisy are solely used for offline computations. The output of our algorithm contains (almost) all the information required for implementation of the verified controller over embedded systems. Therefore, the same savings can be expected when the controller is implemented on an actual embedded system.
\end{remark}}
\begin{comment}
\textcolor{red}{Sadegh: This is long comment on the content of this section. I cannot figure out how this line of reasoning is mapped to the implementation. We can discard my comment given that we are close to the deadline.}

### 2019-07-17 emsoft-2019/camera_ready/intro.tex
colors-->blue

For smaller examples, the analysis takes only a few minutes of computation.
We also demonstrate the scalability of our mixed-precision tuning and error bounds analysis:
we show that on explicit MPC controllers with thousands of regions, our tool finishes in a few hours.
In absolute terms, the memory saving corresponds to 21KB over a uniform
precision implementation on our largest benchmark. %\Mahmoud{It should be noticed that these savings are not affecting the run-time. Synthesis of the controller and the corresponding precisions are all performed offline.}
\Mahmoud{It should be noticed that EMPC is a class of controllers which is mainly appropriate for implementation over processors with limited computational power and memory. Typical memory of these processors is in the order of tens of kilobytes (e.g. $32$ KB) and hence $21$ KB is considered as a significant saving. Also, memory savings are not affecting the run-time as synthesizing of the controller and the corresponding mixed precision computations are all performed offline.}\\
In summary, our contributions are:
\begin{itemize}
\item We provide an end-to-end automatic tuning algorithm for robust explicit MPC that ensures \Mahmoud{that}
hard constraints on the state are met despite numerical errors;
\item We design a static and scalable error analysis algorithm for piecewise affine functions to bound
errors in controller implementations; and
\item Through a set of standard control benchmarks, we demonstrate that our tuning algorithm
can achieve up to 20\% savings in memory while maintaining robustness.
\end{itemize}

>>>>

For smaller examples, the analysis takes only a few minutes of computation.
We also demonstrate the scalability of our mixed-precision tuning and error bounds analysis:
we show that on explicit MPC controllers with thousands of regions, our tool finishes in a few hours.
In absolute terms, the memory saving corresponds to 21KB over a uniform
precision implementation on our largest benchmark. %\Mahmoud{It should be noticed that these savings are not affecting the run-time. Synthesis of the controller and the corresponding precisions are all performed offline.}
\Mahmoudd{It should be noticed that EMPC is a class of controllers which is mainly appropriate for implementation over processors with limited computational power and memory. Typical memory of these processors is in the order of tens of kilobytes (e.g. $32$ KB) and hence $21$ KB is considered as a significant saving. Also, memory savings are not affecting the run-time as synthesizing of the controller and the corresponding mixed precision computations are all performed offline.}\\
In summary, our contributions are:
\begin{itemize}
\item We provide an end-to-end automatic tuning algorithm for robust explicit MPC that ensures \Mahmoud{that}
hard constraints on the state are met despite numerical errors;
\item We design a static and scalable error analysis algorithm for piecewise affine functions to bound
errors in controller implementations; and
\item Through a set of standard control benchmarks, we demonstrate that our tuning algorithm
can achieve up to 20\% savings in memory while maintaining robustness.
\end{itemize}
